---
title: "PANTS Vignette"
author: "Jonathan Dreyfuss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Welcome to the `PANTS` package for Pathway Analysis, and optionally integration, via NeTwork Smoothing. In this vignette, I'll illustrate a case study of the package for integrating proteomics and metabolomics data to identify significant pathways.

## Install
See README.md for installation.

## Load package
```{r lib, message=FALSE}
library("PANTS")
```

## Load interaction network
We'll use the Pathway Commons v9 network in Simple Interaction Format (SIF), which holds interactions in the format: physical_entity_id <relationship type> physical_entity_id. The 1st and 3rd columns represent the edges of the network, called an "edge list". This dataset has been included in `PANTS`, so you can load it using:
```{r pc9}
data("pc9")
```

## Construct graph
`PANTS` smooths statistics over the network using a network kernel derived from a graph. So we transform `pc9` into a simplified graph:
```{r gr}
gr <- igraph::simplify(igraph::graph_from_edgelist(pc9, directed = FALSE))
```

## Construct predefined pathways
We need to load predefined pathways that consist of genes and metabolites whose IDs are consistent with those of our interaction network. One source for these pathways is the Small Molecule Pathway Database (smpdb.ca). SMPDB offers one file for proteins and another for metabolites. If you read these in, you can construct pathways using the `SMPDB2Gmat` function. I've already done this and included the result with the package. You can load this data, and then enforce its rows to match the kernel with:
```{r gmat}
data("smpdb_gmat")
```

## Calculate kernel
```{r ker}
ker <- graph2kernel(gr)
```

## Get data
To illustrate `PANTS`, we'll simulate protein and metabolite abundance data on 10 samples, with 5 cases and 5 controls. The rows (features) of this data matrix need to overlap with the rows of `gmat`. We'll sample features from `gmat` instead of using all the features to save computation time. We also need a vector of phenotypes.
```{r sim}
set.seed(42)
features <- rownames(smpdb_gmat)[sample(x=1:nrow(smpdb_gmat), size=10^3)]
M <- matrix(rexp(n=10^4), nrow=10^3, ncol=10, dimnames=list(features, paste0("s", 1:10)))
pheno.v <- setNames(rep(c("case", "ctrl"), each=5), nm=colnames(M))
```

## Pathway analysis
We execute the main function, `pants`, by testing a contrast (or comparison) on this dataset, e.g. case vs. control. We limit the number of permutations to speed up the calculation. In general, if the `object` or `nperm` is large, this step will be slow.
```{r pants}
contr <- c(CASEvsCTRL="case-ctrl")
res.lst <- pants(object=M, phenotype.v=pheno.v, contrast.v=contr, ker=ker, Gmat=smpdb_gmat, nperm=50)
pwy.stats <- res.lst$pwy.stats
```

## Plot
Draw network diagram of driving nodes for the pathway "Beta-Alanine Metabolism". We annotate select nodes we have annotation for to replace their ChEBI ID with a name.
```{r plot}
score.v <- stats::setNames(res.lst$feature.stats$score, nm=rownames(res.lst$feature.stats))
annot <- c("CHEBI:15846"="NAD+")
plot_pwy(gr=gr, ker=ker, Gmat=smpdb_gmat, pwy="Beta-Alanine Metabolism", score.v=score.v, annot=annot, ntop=5, name=NA)
```
